SPP-Perc: Peceptron Learning Based Enhanced Signature Path Prefetcher

### 2018-11-20 abs.tex
Separating arch and implementation. Filter-Table and Throttler-Filter renaming

In this paper, we introduce Perceptron-based Prefetch Filtering to help make
this prefetching decision accurately. The perceptron layer acts as a throttler
to filter out the unnecessary prefetches recommended by the underlying
prefetcher.  We have also explored a range of features that can be used to
train the perceptron layer.  Our results show that perceptron-based filtering
improves performance on the memory intensive subset of the SPEC 2017 benchmark
suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
to a state-of-the art prefetcher.  We also demonstrate that the performance
gained from using our efficient throttler scales better with increasing numbers
of cores.

>>>>

In this paper, we introduce Perceptron-based Prefetch Filtering to help make
this prefetching decision accurately. The perceptron layer acts as a check 
to filter out the unnecessary prefetches recommended by the underlying
prefetcher.  We have also explored a range of features that can be used to
train the perceptron layer.  Our results show that perceptron-based filtering
improves performance on the memory intensive subset of the SPEC 2017 benchmark
suite by 6.84\% on single-core and by 11.9\% on multi-core traces, as compared
to a state-of-the art prefetcher.  We also demonstrate that the performance
gained from using our efficient filter scales better with increasing numbers
of cores.

### 2018-11-20 back.tex
Separating arch and implementation. Filter-Table and Throttler-Filter renaming

In this section, we compare the existing work done in the domain of
prefetching.  We also discuss interesting memory access patterns seen in the
workloads in SPEC 2017 benchmark.  The second half of this section gives
background for the Signature Path Prefetcher (SPP) architecture.

@@

%Finally, the paper on Data Cache Prefetching with Perceptron
%Learning~\cite{BadPerc} talks about the idea of two step prefetching.
%The first step is an existing baseline prefetcher while the second
%step is a perceptron based throttler.  The paper has lots of
%limitations in terms of design and implementation.  That reflects in
%the fact that it did not lead to any significant performance gain over
%even the basic prefetchers like the Stride Prefetcher~\cite{Stride} and
%the Markov Prefetcher~\cite{Markov}.

\subsection{Perceptron Learning in Architecture}
\label{Background-Perceptron}
Perceptron learning for computer architecture design has been around for a
while.  It was first popularized by Jim\'enez \textit{et. al.} for conditional
branch prediction~\cite{Perc_Branch}. The biggest advantage of the perceptron
model is that it can adjust its weights dynamically to increase / decrease the
importance given to each feature - branch histories in this case.  The
structure we use is derived from the model introduced for Perceptron-based
Reuse Prediction~\cite{Perc_Reuse}, refined with many additional features as
Multiperspective Reuse Prediction{Multiperspective}, and originally derived
from Hashed Perceptron branch prediction{HashedPerceptron}.  Here, a hash of
the feature itself is used to index into a table of perceptron weights.

@@

Perceptron Learning for Reuse Prediction~\cite{Perc_Reuse} extended the hashed
perceptron architecture to perform prediction in context of cache replacement.
The perceptron layer learns to correlate cache replacement behaviour with a
variety of features. The features introduced in that paper were derived from
program counters of the current and the last few memory access instructions
and from the current memory access address.  This led to a prediction
mechanism that is highly accurate, has quick learning rate and is adaptable to
changing program behaviour.

The work was further extended by Multiperspective Reuse
Prediction~\cite{Multiperspective}, which introduced a varied set of
parametric features.  Although design space exploration for feature and
parameter selection was a non-trivial task, once fixed, it yields
replacement predictor highly suitable for the given set of applications.

@@

\textit{Perceptron Update Rule:} In all the above implementations, a uniform
perceptron update principle is followed.  The weights need to be updated if
the prediction was wrong or the magnituded of the dot product does not exceed
a certain threshold.  After a certain training period, these weights are
proportional to the probability of the outcome being true.  This training
threshold makes sure that the perceptrons are trained until a certain level of
confidence and yet they are not over-trained to the given set.  Care is taken
that the weights saturate at the maximum positive and negative values so that
they remain confined to the bit-width.  The same rules will be applied in the
discussion done in Section \ref{Design-Perceptron} on PPF learning algorithm.

\subsection{Prefetching with SPEC 2017} \label{Background-SPEC2017}
% djimenez: we might not need to discuss the benchmarks too much as they are
% standard. we don't have a lot of room anyway.

\textit{<Discuss SPEC 2017 traces here> <What are the possible graphs /
analysis tools we can use for SPEC traces?>}

@@

\textbf{Signature Table:} The ST keeps track of 256 most recently accessed
pages.  It is meant to capture memory access patterns within a page
boundary.  SPP indexes into an entry of ST using the page number.  For
each entry corresponding to a page, ST stores the `last block offset'
and `old signature'.  Last block offset is the block offset of the
last memory access of that given page.  The block offset is calculated
with respect to the page boundary.  The signature is a 12-bit
compressed representation of the past few memory accesses for that
page.  The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
Delta is the numerical difference between the block offset of the 
current and the previous memory acces. In case a matching page entry 
is found, the stored signature retrieved and used to index into the 
Pattern Table.

\textbf{Pattern Table:} The PT is indexed by the signature generated
from ST.  PT holds predicted delta patterns and their confidence
estimates.  Each entry indexed by the signature holds up to 4 unique
delta predictions.  This is implemented by making PT as a 4-way
associative table.

@@

\textbf{Prefetch Filter:} The PF is a 1024 entry 1-way associative table that
keeps a record of last few entries prefetched.  This proves useful for
updating the prefetcher states if a tracked prefetch leads to a demand hit or
a cache eviction.

\textbf{Lookahead Prefetching:} On each trigger, SPP tries to go down
the program speculation path using its own prefetch suggestion.
Using current prefetch as the baseline, it re-accesses the PT to generate further
prefetches.  It keeps on repeating this cycle of accessing PT and
updating the signature based on highest confidence prefetch from the
last iteration.  These iteration count till which SPP manages to
predict prefetch entries in the lookahead manner is characterized as
its `depth'.  While doing this, SPP also keeps compounding the
confidence in each depth.  Thus as depth increases, overall confidence
keeps decreasing.

\textbf{Confidence Tracking}: The PT keeps a track of hits to each
signature through a counter C\textsubscript{sig}.  The number of hits
for a given delta per signature are tracked using a counter
C\textsubscript{delta}.  The confidence for a given delta is
approximated through C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}.  When SPP enters into a lookahead mode, the path
confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;  P\textsubscript{d-1}$$ 
Here $\alpha$ represents the global accuracy, calculated as the ratio of 
the number of prefetches which led to a demand hit to the number of 
prefetches recommended in total. The range of $\alpha$ is [0,1].

`D' is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against prefetching-threshold
(T\textsubscript{p}) to decide whether to prefetch or not.  If yes,
then P\textsubscript{d} is thresholded against a numerically bigger
fill-threshold (T\textsubscript{f}) to decide whether the prefetch
should be sent to L2 Cache (high confidence prefetch) or Last Level
Cache (low confidence prefetch)

@@

\subsection{Case for an On-line Throttler}
\label{Background-Case}
As compared to some of the other state of the art prefetchers [BOP], SPP is
less aggressive.  In the single-core environment, this gives BOP an edge as
there is no resource contention among the different cores.  Hence more
aggressive prefetching is bound to prove beneficial.  As we increase the core
count, we observe that SPP starts outperforming rest of the prefetchers.  This
can be attributed to the fact that each prefetch suggested by SPP is a
carefully calculated one and that prevents cache pollution.  \textit{<Figure
showing the variation of SPP vs BOP wrt core count>} For 4-core applications,
SPP suggests XX\% fewer prefetches than BOP and yet leads to higher IPC.

The above analysis shows that with a more careful throttling mechanism, any
prefetcher like SPP can be tuned to become much more aggressive, leading to
increased coverage.  The onus of maintaining the accuracy now falls on the
independent throttler.  To test the hypothesis, we tuned down SPP to the
minimum possible threshold, with the effect of increasing prefetch suggestions
made by SPP by XX\%.  \textit{<Comparison of SPP-Unleased with SPP / BOP>}
Obviously this came at a cost of increased DRAM traffic and cache pollution.

@@

To that effect, we introduce an independent on-line perceptron based
throttling mechanism.

>>>>

In this section, we compare the existing work done in the domain of
prefetching.  We also give a brief background for the 
Signature Path Prefetcher (SPP) architecture.

@@

%Finally, the paper on Data Cache Prefetching with Perceptron
%Learning~\cite{BadPerc} talks about the idea of two step prefetching.
%The first step is an existing baseline prefetcher while the second
%step is a perceptron based filter.  The paper has lots of
%limitations in terms of design and implementation.  That reflects in
%the fact that it did not lead to any significant performance gain over
%even the basic prefetchers like the Stride Prefetcher~\cite{Stride} and
%the Markov Prefetcher~\cite{Markov}.

\subsection{Perceptron Learning in Architecture}
\label{Background-Perceptron}
Perceptron learning for computer architecture design has been around for a
while.  It was first popularized by Jim\'enez \textit{et. al.} for conditional
branch prediction~\cite{Perc_Branch}. The biggest advantage of the perceptron
model is that it can adjust its weights dynamically to increase / decrease the
importance given to each feature - branch histories in this case.  
Perceptron-based Reuse Prediction~\cite{Perc_Reuse} implemented the hashed
perceptron architecture to perform prediction in context of cache replacement.
The structure  we use is derived from the model introduced in that paper, refined with 
many additional features as in Multiperspective Reuse Prediction{Multiperspective}.  
The model was originally derived from Hashed Perceptron branch 
prediction~\cite{HashedPerceptron}.  Here, a hash of the feature itself is used to 
index into a table of perceptron weights.

@@

@@

\textit{Perceptron Update Rule:} In all the above implementations, a uniform
perceptron update principle is followed.  The weights need to be updated if
the prediction was wrong or the magnituded of the dot product does not exceed
a certain threshold.  After a certain training period, these weights are
proportional to the probability of the outcome being true.  This training
threshold makes sure that the perceptrons are trained until a certain level of
confidence and yet they are not over-trained to the given set.  Care is taken
that the weights saturate at the maximum positive and negative values so that
they remain confined to the bit-width.  The same rules will be applied in the
discussion done in Section \ref{Arch-Perceptron} on PPF learning algorithm.

%\subsection{Prefetching with SPEC 2017} \label{Background-SPEC2017}
% djimenez: we might not need to discuss the benchmarks too much as they are
% standard. we don't have a lot of room anyway.

%\textit{<Discuss SPEC 2017 traces here> <What are the possible graphs /
%analysis tools we can use for SPEC traces?>}

@@

\textbf{Signature Table:} The Signature Table keeps track of 256 most recently accessed
pages.  It is meant to capture memory access patterns within a page
boundary.  SPP indexes into an entry of the Signature Table using the page number.  For
each entry corresponding to a page, the table stores a `last block offset'
and an `old signature'.  Last block offset is the block offset of the
last memory access of that given page.  The block offset is calculated
with respect to the page boundary.  The signature is a 12-bit
compressed representation of the past few memory accesses for that
page.  The signature is calculated as:
$$New Signature = (\,Old Signature << 3 bits\,) \;\;XOR\;\; (\,Delta\,)$$ 
Delta is the numerical difference between the block offset of the 
current and the previous memory acces. In case a matching page entry 
is found, the stored signature retrieved and used to index into the 
Pattern Table.

\textbf{Pattern Table:} The Pattern Table is indexed by the signature generated
from the Signature Table.  Pattern Table holds predicted delta patterns and their confidence
estimates.  Each entry indexed by the signature holds up to 4 unique
delta predictions.  This is implemented by making the Pattern Table as a 4-way
associative table.

@@

\textbf{Prefetch Table:} The Prefetch Table is a 1024 entry 1-way associative table that
keeps a record of last few entries prefetched.  This proves useful for
updating the prefetcher states if a tracked prefetch leads to a demand hit or
a cache eviction. \textit{Note:} The original paper refers to this as the Prefetch Filter 
but since in our context filter refers to the perceptron filter, we will be calling 
this structre as the Prefetch Table to avoid any confusion.

\textbf{Lookahead Prefetching:} On each trigger, SPP tries to go down
the program speculation path using its own prefetch suggestion.
Using current prefetch as the baseline, it re-accesses the Pattern Table to generate further
prefetches.  It keeps on repeating this cycle of accessing Pattern Table and
updating the signature based on highest confidence prefetch from the
last iteration.  These iteration count till which SPP manages to
predict prefetch entries in the lookahead manner is characterized as
its `depth'.  While doing this, SPP also keeps compounding the
confidence in each depth.  Thus as depth increases, overall confidence
keeps decreasing.

\textbf{Confidence Tracking}: The Pattern Table keeps a track of hits to each
signature through a counter C\textsubscript{sig}.  The number of hits
for a given delta per signature are tracked using a counter
C\textsubscript{delta}.  The confidence for a given delta is
approximated through C\textsubscript{d} = C\textsubscript{delta} /
C\textsubscript{sig}.  When SPP enters into a lookahead mode, the path
confidence P\textsubscript{d} is given as:
$$P\textsubscript{d} \;=\; \alpha  \;.\;  C\textsubscript{d}  \;.\;  P\textsubscript{d-1}$$ 
Here $\alpha$ represents the global accuracy, calculated as the ratio of 
the number of prefetches which led to a demand hit to the number of 
prefetches recommended in total. The range of $\alpha$ is [0,1].

`D' is the lookahead depth. For $d = 1$, when SPP is in
non-speculative mode, P\textsubscript{0} can be thought of as 1. 
The final P\textsubscript{d} is thresholded against PF\_THRESHOLD 
(T\textsubscript{p}) to reject the low confidence suggestions and 
then against a numerically bigger FILL\_THRESHOLD (T\textsubscript{f}) to 
decide whether the prefetch should be sent to L2 Cache 
(high confidence prefetch) or Last Level Cache (low confidence prefetch).
The two thresholds were empirically set to 25 and 90 respectively, 
on the scale of 0 to 100.

@@

\subsection{Case for an On-line Filter}
\label{Background-Case}
As compared to some of the other state of the art prefetchers [BOP], SPP is
less aggressive.  In the single-core environment, this gives BOP an edge as
there is no resource contention among the different cores.  Hence more
aggressive prefetching is bound to prove beneficial.  As we increase the core
count, we observe that SPP starts outperforming rest of the prefetchers.  This
can be attributed to the fact that each prefetch suggested by SPP is a
carefully calculated one and that prevents cache pollution.  \textit{<Figure
showing the variation of SPP vs BOP wrt core count>} For 4-core applications,
SPP suggests XX\% fewer prefetches than BOP and yet leads to higher IPC.

The above analysis shows that with a more careful filtering mechanism, any
prefetcher like SPP can be tuned to become much more aggressive, leading to
increased coverage.  The onus of maintaining the accuracy now falls on the
independent filter.  To test the hypothesis, we tuned down SPP to the
minimum possible threshold, with the effect of increasing prefetch suggestions
made by SPP by XX\%.  \textit{<Comparison of SPP-Unleased with SPP / BOP>}
Obviously this came at a cost of increased DRAM traffic and cache pollution.

@@

To that effect, we introduce an independent on-line perceptron based
filtering mechanism.

### 2018-11-20 design_arch.tex
Separating arch and implementation. Filter-Table and Throttler-Filter renaming

\section{PPF Architecture}
\label{Design}

In this section we describe our implementation of PPF.  PPF is a generalised
prefetch throttler and can be adapted to most prefetchers with minimal
modifications.  For a practical implementation, we have used SPP as the base
engine prefetcher.

\subsection{Changes made to original SPP}
\label{Design-Changes}
To modify the SPP design to suit our scheme, the following changes were made:

\begin{itemize}
\item \textbf{Original Thresholds discarded}\newline In the original
  SPP design, the internal confidence was compared against a preset
  PF\_THRESHOLD (T\textsubscript{p}) to decide whether to accept or reject the
  prefetch suggestion.  Furthermore, the accepted suggestions were thresholded
  against FILL \_THRESHOLD (T\textsubscript{f}) to decide the fill level: L2
  Cache vs Last Level Cache.  T\textsubscript{p} and T\textsubscript{f} were
  set to 25 and 90 respectively, on the scale of 0 to 100.  In PPF, the
  perceptron sum is used to make these decisions.  Hence these two thresholds
  are no longer needed

%%[NOTE] Can this point be removed?
% djimenez: I think this would be necessary to reproduce our results so leave it in.

\item \textbf{Looking at L2 MSHR Queue while prefetching}\newline The original
SPP would not consider the slots available in the L2 MSHR queue.  It had an
assumption that at no time can the number of prefetches suggested by SPP
exceed the capacity of the queue.  While it was not proven in the paper, the
internal confidence mechanism along with the thresholds made sure that the
above assumption was maintained.  In PPF since there can be no assumption on
the degree of aggressiveness of the underlying prefetcher, we explicitly check
that at no point should the number of suggested prefetches exceed the L2 MSHR
queue.

\item \textbf{Enhanced Filter}\newline The original SPP introduces a concept of
the Prefetch Filter that keeps a track of up to 1024 L2 cache prefetch
suggestions.  For PPF, the Filter was modified to store metadata such as program
counter, memory access address etc.  These features are needed to train the perceptron
when the result of that particular prefetch becomes available.  At training time,
this data is used to retrieve the state of the prefetcher when the prediction
was made.

\item \textbf{Reject Filter introduced}\newline The Prefetch Filter mentioned
above keeps track of the prefetches that actually happened in the L2 Cache.
Correspondingly, we introduced a reject filter.  It keeps track of the
prefetches that were suggested by the base SPP engine but were rejected by the
perceptron.  In case it is detected that a future demand fetch could have used
this rejected prefetch, the perceptron weights are updated to reflect that
misprediction.

\subsection{The Perceptron Layer}
\label{Design-Perceptron}
As discussed earlier, the perceptron layer was introduced to act as a
throttler to the underlying prefetch engine.  This section explores the
details of the perceptron layer and the range of features that are ultimately
used by the perceptron for correlating the prefetches.

The weights are stored in separate tables for each
feature.  Features are indexed using different numbers
of bits.  At most 12 bits of a feature are used to index into a
weights table.  Certain features require more resolution power
\textit{i.e.}, the full 12 bits of indexing.  On the other hand, some
features requiring lower resolution require as few as 7 bits of
indexing.  The variable indexing was determined by studying the
features and fine-tuned empirically so as to achieve a good accuracy
vs area trade-off.  Exact details are mentioned under ``Area Overheads''
in Section~\ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15.  Each feature
has its own dedicated table.  Our proposed design uses 9 features.
As program execution begins, all the weights are initialized to 0.

\textbf{Inference}\newline In the ChampSim simulator, the prefetching method
is triggered on every L2 Cache demand access.  At that point, the prefetcher
has the option to do or not do a prefetch -- if it does, then it has a choice
in how many cache lines to prefetch.  The suggested prefetches can be either
in the L2 Cache or the last-level cache.  The prefetcher decides this
placement, and it is usually done based on an internal confidence mechanism.

When the base prefetcher is triggered, it starts suggesting candidates for
prefetching.  All these recommended prefetches are tested in the perceptron
throttler before they can access the Prefetch Filter, as in the original SPP
datapath.

@@

Once all the weights are retrieved, they are summed.  The sum is thresholded
based on the preset threshold PERC\_THRESHOLD\_LO.  Only the prefetch
candidates with perceptron sum higher that the threshold qualify for
prefetching.  The prefetches that qualify through the perceptron throttler
stage go through the Prefetch Filter stage.  The Prefetch Filter eliminates
any redundant prefetch suggestions.  In PPF, all the metadata that is required
to recreate the features is also stored in the Prefetch Filter.  At a later
stage, when the feedback of the current prefetched line is available, the
stored data is used to train the perceptron.

@@

In addition to the Prefetch Filter mentioned above, PPF also maintains a
``Reject Filter.''  The reject filter is a 1024-entry deep direct-mapped
table.  If a prefetch suggestion is rejected by perceptron throttler, it is
logged into the Reject Filter.  The filer is used to train the perceptron to
avoid false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

@@

%As previously stated all the metadata describing the program state at the
%instant of prefetching is stored in the Prefetch Filter or the Reject Filter.
%This information is useful when the perceptron needs to be updated
%subsequently.

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 Cache.  This action triggers training of the base prefetcher as well as the
perceptron throttler.  Training the perceptron throttler involves restoring
the metadata that was stored in Prefetch Filter or the Reject Filter.  This is
done by indexing into the corresponding filters using the cache line address
of the block being used to train. (10-bits indexing + 6-bits tag matching).
Once the state of the program at the time of prefetching is available, it is
used to index into the perceptron weights table.

If a demand access block that triggers the training was tagged as a valid
prefetch in the Prefetch Filter, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 if the
predicted sum does not cross a pre-defined threshold.  If a cache block
eviction led to training and the corresponding valid entry was found in the
Prefetch Filter, then the prediction made by the perceptron was wrong.  The
perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to reflect the
misprediction. In either case, weights are saturated at -16 or +15.

A secondary training mechanism also kicks in during demand fetches.  Before
the demand access triggers the next set of prefetches, Reject Filter is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the underlying prefetch engine but rejected by the
perceptron throttler.  Thus, the perceptron should have been more confident
about that particular prefetch.  Once such a scenario is identified, the state
of the execution at the time of prefetch is retrieved from the Reject Table.
The retrieved data is used to index into the various weights tables of the
perceptron and the corresponding values are updated by +1, saturating between
and -16 and +15.  This update reflects increased confidence for the prediction
corresponding to that prefetch.

@@

\subsection{Features used by Perceptron}
\label{Design-Features}
Here we discuss the various features that correlate the prefetching decision
with the program behaviour.

\begin{itemize}
\item \textbf{Base Address} of the demand access that triggered the
  prefetch.  Since prefetches are triggered from the L2 demand access,
  they tend to be correlated with the triggering address.

\item \textbf{Cache Line} and \textbf{Page Address}: These two
  separate features are derived from the base address that triggered
  the prefetch, as follows: base\_addr >> LOG2\_BLOCK \_SIZE and
  base\_addr >> LOG2\_PAGE\_ SIZE respectively.  The idea behind using
  three different shifted versions of the same feature is that it
  allows us to look into a wider range of bits than with a single
  version.  It also helps give more importance to the overlapping bits
  and lesser importance to most and least significance bits.  This
  approach can also eliminate destructive interference that can be
  caused by directly folding the bits into half.

% djimenez: this is a nice intuition and explanation.

\item \textbf{Program Counter XOR Depth}: The PC is for the instruction
  that triggered the prefetch chain.  Depth refers to the iteration count of
  the lookahead stages.  In general, the PC is not considered as a good basis
  for doing lookahead prefetching as all the prefetches with depth >= 1 are
  aliased into the same PC which would not be true in an actual demand access.
  This feature resolves a PC into a different value for each lookahead depth
  of prefetch speculation, giving a more accurate correlation in lookahead
  cases.  This is akin to the concept of Virtual Program Counters~\cite{VPC}
  introduced by Kim \textit{et. al.} for indirect branch prediction.

\item \textbf{PC\textsubscript{1} XOR PC\textsubscript{2}>>1 XOR
  PC\textsubscript{3}>>2}: Here $PC_i$ refers to the last $i^{th}$
  PC before the instruction that triggered the current prefetch.
  Hashing together the last three PCs tell PPF about the path
  that led to the current demand access and helps capture and
  branching information of the current basic block.  PCs are shifted
  in the increasing order of history before being hashed together.
  This is done to avoid the resultant value of zero in case 2 or more
  PCs are the same.  Additionally, obfuscating the information as it
  gets old allows us to get a wider and an approximate look into the
  program history.

\item \textbf{Program Counter XOR Delta}: This feature tells us if a given PC
  favours particular value(s) of delta.  As noted earlier, while the PC in
  itself does not convey much useful information, this hash resolves the PC
  into different values based on the tendency of that PC to favour a certain
  delta.  Thus, the dynamic nature of different instances of the same memory
  access instruction can be captured here.

\item \textbf{Confidence}: The integer confidence on a scale of 0 to
  100 that was used in the original SPP design.  While the original confidence
  is not used directly for decision making, it can still contribute to the
  final decision made.

% djimenez: this is a little vague. can you be more precise?
\item \textbf{Current Signature XOR Delta}: Recall from
  the discussion of SPP in Section \ref{Background-SPP} that the new signature
  is generated using the old signature and the block delta.  While ``Current
  Signature XOR Delta'' is not the exact formula for generating the future
  signature, it gives an approximate idea of the path that the combination of
  these two values can lead to.

\item \textbf{Page Address XOR Confidence}: This feature scores the tendency
of each page to be prefetch friendly or prefetch averse. It helps resolve a
page into different entries depending on its confidence for prefetching, which
can vary during phases of a program execution.

%%%%%%% Below are the features rejected when downsizing 14 -> 9
% \item \textbf{Current Signature}: The 12-bit signature used by SPP
%   to index into the Pattern Table
% \item \textbf{Delta}: The signed version of the difference in the
%   block address that triggers the prefetch and the block address of
%   the data being prefetched
% \item \textbf{PC}: Program Counter of the Load instruction that
%   triggers that particular prefetch
% \item \textbf{Base address XOR Delta}: If any particular value(s) of
%   delta are highly favoured by a given address which triggered the
%   prefetch, this composite feature can capture that information
% \item \textbf{Cache line XOR Delta}: qThis feature captures if any
%   particular delta is highly favoured by a given cache line address
%\item \textbf{Confidence XOR Delta}: This gives a combined insight
%  into that fact that certain delta-confidence pairqs predicted by
%  SPP might correlate more with the prefetch outcome
%%%%%%%%%%%%%%%%%%%%%%

Some of the composite features are derived from simple hashing (XOR) of two
primary features.  There is always a question of usefulness of such composite
features and the new information added.  We justify the choice of each feature
by quantifying the contribution made towards predicting prefetch behaviour, in
Section \ref{Method-Features}.  Finally, as noted above, each feature indexes
into its independent entry of perceptron weights.

\subsection{Generalizing PPF for any Prefetcher}
\label{Design-Generalizing}
The above implementation of PPF shows that it is highly modular and can be
adapted to be used over any base prefecher for increased prefetch accuracy.
In our implementation, only two hooks are required between PPF and the
baseline SPP. The first is to make sure that all the prefetch candidates of
SPP pass through the perceptron throttler and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

@@

\item \textbf{Feature Selection:} Six out of the nine features we developed
use information derived from program execution, irrespective of the baseline
SPP prefetcher. Beyond that, the feature set can be expanded to convey any
useful information from the prefetcher to the perceptron throttler.  For
example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
and ``Page Address XOR Confidence'' are such features.

>>>>

\section{PPF Architecture}
\label{Arch}

In this section, we talk about the the general structure of the 
Perceptron-Based Prefetch Filtering (PPF) for an underlying prefetcher.  
PPF is a generalised prefetch filtering mechanism and can be adapted to 
most prefetchers with minimal modifications.

\subsection{The Perceptron Filter}
\label{Arch-Perceptron}
The Perceptron Filter is organised as tables of perceptron weights, 
with a separate table for each feature.  
The feature value is used to index into the corresponding table.
This indexing is done using different numbers of bits for each feature.  
At most 12 bits of a feature are used to index into the
weights table.  Certain features require more resolution power
\textit{i.e.}, the full 12 bits of indexing.  On the other hand, some
features requiring lower resolution require as few as 7 bits of
indexing.  The variable indexing was determined by studying the
features and fine-tuned empirically so as to achieve a good accuracy
vs area trade-off.  Exact details are mentioned under ``Area Overheads''
in Section~\ref{Method-Overheads}

A single entry in the table corresponds to a perceptron weight.  Each
weight is a 5 bit counter - saturating at -16 and +15.  
Our proposed design uses nine features, hence there are nine perceptron 
weight tables. As program execution begins, all the weights are initialized to 0.

\textbf{Inference}\newline In the ChampSim simulator, the prefetching method is 
invoked on every L2 Cache demand access.  At that point, the base prefetcher
has the option to do or not do a prefetch -- if it does, then it has a choice
in how many cache lines to prefetch.  The suggested prefetches can be either
in the L2 Cache or the last-level cache.  The prefetcher decides this
placement, and it is usually done based on an internal confidence mechanism.

When the base prefetcher is triggered, it starts suggesting candidates for
prefetching.  All these recommended prefetches have to be tested through the perceptron
filter to qualify for final prefetching.

@@

Once all the weights are retrieved, they are summed.  The sum is thresholded
based on the preset threshold PERC\_ THRESHOLD\_LO.  Only the prefetch
candidates with perceptron sum higher that the threshold qualify for
prefetching.  The prefetches that qualify through the perceptron 
stage are recorded in the Prefetch Table.  In PPF, all the metadata that is required
to recreate the features is also stored in the Prefetch Table.  At a later
stage, when the feedback of the current prefetched line is available, the
stored data can be used to train the perceptron.

@@

In addition to the Prefetch Table mentioned above, PPF also maintains a
``Reject Table.''  The Reject Table is a 1024-entry deep direct-mapped
table.  If a prefetch suggestion is rejected by the perceptron layer, it is
logged into the Reject Table.  The filer is used to train the perceptron to
avoid false negatives \textit{i.e.}, cases where prediction was to reject the
prefetch but the prefetch would have been useful.

@@

%As previously stated all the metadata describing the program state at the
%instant of prefetching is stored in the Prefetch Table or the Reject Table.
%This information is useful when the perceptron needs to be updated
%subsequently.

\textbf{Training}\newline In the prefetching environment, feedback for a
prefetch is received whenever there is an eviction or a demand access from the
L2 Cache.  This action triggers training of the base prefetcher as well as the
perceptron layer.  Training the perceptrons involves restoring
the metadata that was stored in Prefetch Table or the Reject Table.  This is
done by indexing into the corresponding tables using the cache line address
of the block being used to train. (10-bits indexing + 6-bits tag matching).
Once the state of the program at the time of prefetching is available, it is
used to index into the perceptron weights table.

If a demand access block that triggers the training was tagged as a valid
prefetch in the Prefetch Table, then the earlier prefetch prediction was
correct.  In that case the perceptron weights are incremented by 1 
if the predicted sum does not cross a pre-defined threshold. These training 
thresholds are introduced to limit the degree of update on correct predictions.
Doing this helps avoid overfitting of the perceptron weights to the given program 
behaviour. These thresholds are referred to as POS\_UPDATE\_THRESHOLD ($\theta_p$) and 
NEG\_UPDATE\_THRESHOLD ($\theta_n$),  respectively for the positive and 
negative values of training saturation.

If a cache block
eviction led to training and the corresponding valid entry was found in the
Prefetch Table, then the prediction made by the perceptron was wrong.  The
perceptron should have ideally rejected the prefetch suggestion as a
low-confidence prefetch.  Here the weights are decremented by 1 to reflect the
misprediction. In either case, weights are saturated at -16 or +15.

A secondary training mechanism also kicks in during demand fetches.  Before
the demand access triggers the next set of prefetches, Reject Table is
checked for a valid entry.  A hit means that the corresponding cache line was
initially suggested by the underlying prefetch engine but rejected by the
perceptron filter.  Thus, the perceptron should have been more confident
about that particular prefetch.  Once such a scenario is identified, the state
of the execution at the time of prefetch is retrieved from the Reject Table.
The retrieved data is used to index into the various weights tables of the
perceptron and the corresponding values are updated by +1, saturating between
and -16 and +15.  This update reflects increased confidence for the prediction
corresponding to that prefetch.

@@

\subsection{Generalizing PPF for any Prefetcher}
\label{Arch-Generalizing}
The above discussion of PPF shows that it is highly modular and can be
adapted to be used over any base prefecher for increased prefetch accuracy.
In general, only two hooks are required between PPF and the
baseline prefetcher. The first is to make sure that all the prefetch candidates of
the prefetcher pass through the perceptron filter and if qualified, the metadata for
perceptron indexing be stored. The second is needed when the feedback of a
prior prefetch is available in form of a subsequent demand hit or cache
eviction. In that case, the stored metadata needs to be retrieved to update
the state of the perceptrons.

@@

\item \textbf{Feature Selection:} In Section~\ref{Impl-Features}, we show the features 
used for PPF. Six out of the nine features we developed
use information derived directly from program execution, irrespective of the baseline 
prefetcher. Beyond that, the feature set can be expanded to convey any
useful information from the prefetcher to the perceptron layer.  The 
methodology explained in Section~\ref{Method-Features} talks about developing a minimal 
feature set for PPF.
%For example, in our implementation ``Confidence,'' ``Current Signature XOR Delta,''
%and ``Page Address XOR Confidence'' are such features.
